---
title: "BMI551hw5"
author: "MK"
date: "Monday, February 29, 2016"
output: html_document
---

#Q1.a.: construct transmission matrix
```{r}
TM <- matrix(c(c(0.15, 0.85),c(0.2, 0.8)), ncol=2, byrow=TRUE)
rownames(TM) <- c('AT-rich', 'GC-rich') ; colnames(TM) <- c('AT-rich', 'GC-rich')
```

#Q1.b.: construct emmision matrix
```{r}
EM <- matrix(c(c(0.3, 0.1, 0.2, 0.4),c(0.05, 0.3, 0.55, 0.1)), nrow=2, ncol=4, byrow=TRUE)
rownames(EM) <- c('AT-rich', 'GC-rich') ; colnames(EM) <- c('A','C','G','T') 
```

#Q1.c.: generate a 100bp DNA seq by employing an HMM built from the transition mat, emission mat, start probabilities (assume equal), and number of bp for the newly generated sequence
```{r}
#using the example from BMI551_Mahyari_ClassDemo as a reference
#Implement a HMM simulation function to generate 100bp DNA sequence
HMMsim = function(trans, emmiss, start, N){
  #nucleotide alphabet
  nucs <- c('A','C','G','T')
  #AT rich or GC rich states
  state_list <- c('AT-rich', 'GC-rich')
  #initialize list to store the prev states for each nucleotide 
  #all_states<-list() returned error so character()
  all_states <- character()
  #initialize list to store the generated 100bp sequence
  seq_list <- character()
  #randomly choose state for the first generated nucleotide (50:50 chance)
  currentState <- sample(state_list, 1, prob=start)
  cat("**** Starting on state:", currentState, "\n**********\n")
  #use the emmision mat to look up the probabilities associated with each nucleotide for the 
  #current state and randomly choose a nucleotide
  nuc <- sample(nucs, 1, prob=emmiss[currentState,])
  cat("**** Starting nucleotide:", nuc, "\n**********\n")
  #keep track of the state that generated the first nucleotide
  all_states[1] <- currentState
  #generate nucleotide sequence
  seq_list[1] <- nuc
  
  # Make N more steps through the HMM to generate the rest of the sequence
  for (i in 2:N)
  {
    #look up the state that generated the previous nucleotide 
    prevState <- all_states[i-1]
    #use the transition mat to look up the probabilities associated with the prevState and randomly
    #choose the new state
    newState <- sample(state_list, 1, prob=trans[prevState,])
    #use the emmision mat to look up the probabilities associated with each nucleotide for the 
    #newState and randomly choose the ith nucleotide
    nuc <- sample(nucs, 1, prob=emmiss[newState,])
    cat(i, " ***", prevState, "->", newState, "Nucleotide:", nuc, "\n")
    #store current ith state
    all_states[i] <- newState
    #store current ith nuc
    seq_list[i] <- nuc
  }
  cat("**** Ending on state:", all_states[N], "\n******************\n");
  #return the sequence and states
  return (list(seq = seq_list,states = all_states))
}
```
#run the HMMsim sequence generator function with params: TM, EM, 50:50 start probability, N=100bp
```{r}
test_seq <- HMMsim(TM, EM, c(0.5,0.5), 100)
#print the generated sequence i.e. ith nucleotide and the ith state
for (i in 1:length(test_seq$seq))
{
  print(paste('Nuc: ',test_seq$seq[i], ' State: ', test_seq$states[i]))
}
```

#Q1.d. : Infer the AT or GC rich states given a DNA sequence i.e. use our HMM to find the most probable state path
#As discussed in class, we must implement the VITERBI ALGORITHM!! using Shannon's lecture from last year (HMM_2015.pdf) as reference for the Viterbi algorithm as well as this R textbook (https://cran.r-project.org/doc/contrib/Krijnen-IntroBioInfStatistics.pdf)
#implement a Viterbi function to calculate the most probable state path
```{r}
#The first step is to construct a Viterbi matrix given our sequence, transition mat, and emmission mat
Viterbi <- function(DNA_seq, trans, emiss)
{
  #construct a matrix with number of nucleotides as rows and number of states as cols
  VM <- matrix(NA, nrow=length(DNA_seq), ncol=dim(trans)[1])
  #step 1: initialize the matrix i.e. Vbegin(0)=1  Vnotbegin(0)=0
  #fill first row with zeros noting that each row represents the ith index in the DNA_seq
  VM[1,] = 0
  #fill in the beggining 
  VM[1,1] = 1
  #populate the VM matrix i.e. calculate most likely states iteratively through observed DNA_seq
  #iterate through each row of the matrix i.e. each nucleotide position in the given sequence
  for (i in 2:length(DNA_seq))
  {
    #consider each state's probability given the observed nucleotide
    for (st in 1:dim(trans)[1])
    {
      #using the emmission mat we consider each nucleotide and calculate the probability of the  
      #current state index, st, emmiting the ith nucleotide
      nuc_prob = emiss[st,DNA_seq[i]]
      #fill out the cell of the matrix with the product of the current state probability and the max
      #of the previous position v[(i-1),] and the transition probability given the current state,
      #st, using the transition mat
      #step 2
      VM[i,st] <- nuc_prob*max(VM[(i-1),]*trans[,st])
      #e.g. el(xi+1)*max{vk(i)*akl}
    }
  }
  #Find the most probable state path by iterating through each row of the VM matrix, i.e. each nucleotide
  #in the sequence, and compute which state has the max probability for that nucleotide
  #step 3
  path = numeric()
  for (i in 1:nrow(VM))
  {
    #?which.max
    path[i] <- which.max(VM[i,])
  }
  
  #return the most probable state path!
  return (path)
}
```
#run the Viterbi algorithm given the observed sequence, we return the most probable state path
```{r}
seq <- c("A", "A", "G", "C", "G", "T", "G", "G", "G", "G", "C", "C", "C", "C", "G", "G", "C", "G", "A", "C", "A", "T", "G", "G", "G", "G", "T", "G", "T", "C")
v_test <- Viterbi(seq,TM,EM)
#print the most probable states i.e. ith nucleotide and the ith state
for (i in 1:length(v_test))
{
  if (v_test[i] == 1){v_test[i]=rownames(EM)[1]}
  else {v_test[i]=rownames(EM)[2]}
  print(paste('Nuc: ',seq[i], ' State: ', v_test[i]))
}
```